Loan Data from Prosper by Emily Lau
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

# install.packages('ggplot2', dependencies = T)
# install.packages("dplyr")
library(ggplot2)
library(dplyr)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
loanData <- read.csv('prosperLoanData.csv')
```

Prosper Marketplace, Inc. is a San Francisco, California-based company in the peer-to-peer lending industry. Prosper Funding LLC, one of its subsidiaries, operates Prosper.com, a website where individuals can either invest in personal loans or request to borrow money. This data set contains 113,937 loans with 81 variables on each loan, including loan amount, borrower rate (or interest rate), current loan status, borrower income, borrower employment status, borrower credit history, and the latest payment information.

# Univariate Plots Section

### Data Structure
```{r echo=FALSE, structure}
str(loanData)

```

### Summary of the dataset
```{r echo=FALSE, summary}
summary(loanData)

```

### 1. Loan Status
Let’s take a look at what are the different loan status and which status are most borrower at?
```{r echo=FALSE, loanStatus}
table(loanData$LoanStatus)

```

**There are 12 different loan status:**

1. Cancelled
2. Chargedoff
3. Completed
4. Current
5. Defaulted
6. FinalPaymentInProgress
7. Past Due (>120 days)
8. Past Due (1-15 days)
9. Past Due (16-30 days)
10. Past Due (31-60 days)
11. Past Due (61-90 days)
12. Past Due (91-120 days)

* There are 6 different kinds of `PastDue` classified by different number of due days. Since I'm only concern about whether a loan is under high risk or not, I have combined all loans under 121 days of `PastDue` into one variable: `PastDue`.
```{r echo=FALSE, combinedPastDue}

#create list that I want to combine into PastDue variable
PastDue <- c("Past Due (1-15 days)",
             "Past Due (16-30 days)",
             "Past Due (31-60 days)",
             "Past Due (61-90 days)",
             "Past Due (91-120 days)")
```

* Based on the Prosper website, a borrower loan is charged-off when it reaches 121 days past due unless the borrower is otherwise notified by Prosper. There are 16 obseravtion with `Past Due(> 120 days)` so I will combined them into `Chargedoff` variable.
```{r echo=FALSE, combinedChargedoff}
#create list that I want to combine into Chargedoff variable
Chargedoff <- c("Past Due (>120 days)")
```

* There are 205 observations under `FinalPaymentInProgress` and since it is almost equivalent to `Completed`, I will combined it with `Completed`. 
```{r echo=FALSE, combinedCompleted}
#create list that I want to combine into Completed variable
Completed <- c("FinalPaymentInProgress")
```

```{r echo=FALSE, updateChanges}
#change LoanStatus format to string
loanData$LoanStatus <- as.character(loanData$LoanStatus)
# start change each variable
loanData$LoanStatus[loanData$LoanStatus %in% Chargedoff] <- "Chargedoff"
loanData$LoanStatus[loanData$LoanStatus %in% PastDue] <- "PastDue"
loanData$LoanStatus[loanData$LoanStatus %in% Completed] <- "Completed"
loanData$LoanStatus <- factor(loanData$LoanStatus, 
                          levels = c("Completed",  "Current", 
                                     "PastDue", "Chargedoff", "Defaulted", "Cancelled"), 
                          ordered = T)
# print out the change result
table(loanData$LoanStatus)
```

The bar chart of the loan status:

```{r echo=FALSE, LoanStatusPlot, dpi=36, out.width="700px", out.height="700px"}
# plot LoanStatus bar chart
ggplot(aes(x= LoanStatus, y = ..count../sum(..count..)), data = loanData) + 
  geom_bar() +
  xlab('Loan Status') +
  ylab('Proportion of Borrowers') +  
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +
      ggtitle("Distribution of Loan Status")
```

**Observations:**

* 34% of the loans are completed.
* The proportion of current loan is about 50%, which means that we are not sure if half of the loan data are defaulted or completed.
* 15% of the loans are chargedoff and defaulted which means there is risk that the investor may loss money.


### 2. Occupation

What are the borrower's occupation?

```{r echo=FALSE, occupation}
table(loanData$Occupation)

```

The bar chart of the borrower's occupation:
```{r echo=FALSE, top10Ocupation}
# plot Employment Status bar chart
ggplot(aes( x = Occupation ), data = subset(loanData, Occupation != '')) +
  geom_bar(fill = '#599ad3' ) +
  theme(axis.text.x = element_text(angle = 90, hjust=0.95,vjust=0.2, size =10)) +
        ggtitle("Distribution of Loans by occupation of the borrower")
```

```{r echo=FALSE, occupationPercentage}
# get borrower occupation percentage
OccupationPercentage <- table(loanData$Occupation)
round(prop.table(addmargins(OccupationPercentage)),digits=2)
```

**Observations:**

* There are a total of 68 different occupations.
* Most of the borrower's occupation is `Others`(13%) and `Professional` (6%). Seems like most borrower did not list down their occupation.


### 3. Employment Status

What are the different employment status?

```{r echo=FALSE, employmentStatus}
table(loanData$EmploymentStatus)
```

**There are 9 different employment status:**

1. Empty
2. Employed
3. Full-time
4. Not available
5. Not employed
6. Other
7. Part-time
8. Retired
9. Self-employed

* There are 2,255 empty status and 5,347 `Not available` status. I will hide them for the plot since they are not required.
* `Other` and `Not available` seems similar so that will be combined as `Other`.
```{r echo=FALSE, employmentStatusClean}
# combine both not available with other
Other <- c("Not available", 
           "Other")

loanData$EmploymentStatus[loanData$EmploymentStatus %in% Other] <- "Other"

# remove empty and Not available employment status
cleanEmploymentStatus <- loanData[!(is.na(loanData$EmploymentStatus) | 
                                      loanData$EmploymentStatus == ""), ]
```

The bar chart of the employment status:
```{r echo=FALSE, employmentStatusPlot_Cleaned}
# plot Employment Status bar chart
ggplot(data=cleanEmploymentStatus, aes(x= EmploymentStatus), y = ..count../sum(..count..)) + 
  geom_bar() +
  xlab('Employment Status') +
  ylab('Proportion of Borrowers') +  
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +
      ggtitle("Distribution of Borrower's Employment Status")
```

**Observations:**

* Quite a number of employment status are left empty (2,255).
* Most of the borrowers are with stable income. 63% of the borrowers are employed and 25% of the borrowers are employed full-time.
* 8.2% of the borrowers are with Other employment status.

### 4. Is Borrower Homeowner

Are the borrower mostly homeowner or not?

```{r echo=FALSE, isBorrowerHomeowner}
table(loanData$IsBorrowerHomeowner)
```

The bar chart of whether borrow is homeowener:
```{r echo=FALSE, IsHomeownerPlot, dpi=36, out.width="700px", out.height="700px"}
# plot LoanStatus bar chart
ggplot(aes(x= IsBorrowerHomeowner, y = ..count../sum(..count..)), data = loanData) + 
  geom_bar() +
  xlab('Is Borrower Homeowner') +
  ylab('Proportion of Borrowers') +  
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +
      ggtitle("Propotion of homeowner")
```

**Observations:**

* There is slightly more borrower who are homeowner (50.5%) than those who are not a homeowner (49.6%).


### 5. Income Range

What are the income range of the borrowers?

```{r echo=FALSE, incomeRange}
table(loanData$IncomeRange)

```

* `Not displayed` values will be hidden since it is not useful for our analysis.
```{r echo=FALSE, rearrangeOrder}
# remove empty and Not available employment status
cleanIncomeRange <- loanData[!(is.na(loanData$IncomeRange) | 
                                      loanData$IncomeRange == "" | 
                                 loanData$IncomeRange == "Not displayed"), ]
```

The bar chart on Income Range:
```{r echo=FALSE, IncomeRangePlot, dpi=36, out.width="1000px", out.height="700px"}
# plot Income Range bar chart
ggplot(aes(x= IncomeRange, y = ..count../sum(..count..)), data = cleanIncomeRange) + 
  geom_bar() +
  xlab('IncomeRange') +
  ylab('Proportion of Borrowers') +  
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +
      ggtitle("Distribution of Borrower's Income Range")
```

**Observations:**

* Seems like the number of borrowers who are `Not employed`(835) under `Income Range` is different from the number of borrowers who are `Not employed` under `Employment Status`(806).


### 6. Borrower State

Where are most borrower from?

```{r echo=FALSE, borrowerState}
table(loanData$BorrowerState)

```

```{r echo=FALSE, borrowerStatePlot}
ggplot(aes( x = BorrowerState ), data = subset(loanData, BorrowerState != '')) +
  geom_bar(fill = '#599ad3' ) +
  theme(axis.text.x = element_text(angle = 90, hjust = .5, vjust = 0.5)) +
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +

          ggtitle("Distribution of Loans by state of the borrower")
```

**Observations:**

* Most of the borrowers are from California (13.6%) followed by Texas (6.3%) and New York (6.21%).


### 7. Loan Term

Let’s take a look at the loan term to see how long borrowers usually opt for loan?

```{r echo=FALSE, loanTerm}
table(loanData$Term)

```


The bar chart on loan term:
```{r echo=FALSE, loanTermPlot, dpi=36, out.width="1000px", out.height="700px"}
# plot Income Range bar chart
ggplot(aes(x= (Term/12), y = ..count../sum(..count..)), data = loanData) + 
  geom_bar() +
  xlab('Loan Term (years)') +
  ylab('Proportion of Borrowers') +  
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +
        ggtitle("Distribution of Loan Term")
```

**Observations:**

* There are 3 types of loan term - 12 months, 36 months and 60 months.
* 77% of the borrowers opt for 3 years loan term.
* Only 1% of the borrowers opt for 1 year loan term. Seems like most borrower are not interested in short loan term.


### 8. Prosper Rating (Alpha)

Let’s take a look at the prosper rating (Alpha)

```{r echo=FALSE, prosperRating}
table(loanData$ProsperRating..Alpha.)

```

```{r, Prosper Rating}
# plot LoanStatus bar chart
loanData$ProsperRating <- factor(loanData$ProsperRating, 
                          levels = c('AA', 'A', 'B', 'C', 'D', 'E', 'HR', 'NA'), 
                          ordered = T)

ggplot(aes(x= ProsperRating, y = ..count../sum(..count..)), data = loanData) + 
  geom_bar() +
  xlab('Prosper Rating') +
  ylab('Proportion of Borrowers') +  
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +
      ggtitle("Distribution of Prosper Rating")
```

**Observations:**

* There is huge percentage of prosper rating as NA (25.53%). 
* The highest prosper rating is C with 16.1% followed by B(13.68%) and A(12.77%).














# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

Our dataset consists of 81 variables, with 11,3937 observations.

### What is/are the main feature(s) of interest in your dataset?

Borrower: `Prosper Score` and `Proser Rating` seems to be the main indicators of a quality of borrowers.

Lender: `Lender Yield` seems to the most important factor for investor.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Properties about borrower such as `Employment Status`, `Income Range`, `Loan Term`.

### Did you create any new variables from existing variables in the dataset?

Yes I created some new variables and factorized a few others. Also the date into month and year for analysis.


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

### 1. Lender Yield vs Prosper Rating

Is there a relation between lender yield and prosper rating?

```{r echo=FALSE, ProsperRatingvsLenderYield}
loanData$ProsperRating..Alpha. <- factor(loanData$ProsperRating..Alpha., levels = c("AA", "A", "B", "C", "D", "E", "HR") )
# create boxplot
ggplot(aes( x = ProsperRating..Alpha., y = LenderYield ), data = subset(loanData, ProsperRating..Alpha. != "")) +
  geom_boxplot() + 
    ggtitle("Lender Yield vs Prosper Rating")
```

**Observation:**

* There's a clear relation between the two, the higher the lender yield, the poorer is the Prosper Rating.

### 2. Borrower Rate vs Prosper Rating

Is there a relation between borrower rate and prosper rating?

```{r echo=FALSE, borrowerRatevsProsperRating}
ggplot(aes(x = ProsperRating..Alpha., y = BorrowerRate), data = subset(loanData, ProsperRating..Alpha. != "")) +
  geom_boxplot() +
    ggtitle("Borrower Rate vs Prosper Rating")

```

**Observation:**

* The borrower rate actually is quite similar to the Lender Yield. The relationship observed here is also similar to the lender yield.


### 3. Loan Orignal Amount vs Prosper Rating

Is there a relation between loan original amount and prosper rating?

```{r echo=FALSE, Bivariate_Plots}
ggplot(aes( x = ProsperRating..Alpha., y = LoanOriginalAmount ), data = subset(loanData, ProsperRating..Alpha. != "")) +
  geom_boxplot()  + 
    coord_cartesian(ylim = c(0,20000)) + 
      ggtitle("Loan Orignal Amount vs Prosper Rating")

```

**Observation:**

* The Median Loan Amount appears to decrease as the prosper rating gets poorer.



```{r echo=FALSE, Bivariate_Plots}


```








# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!