Loan Data from Prosper by Emily Lau
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# install.packages('ggplot2', dependencies = T)
# install.packages("dplyr")
# install.packages("lubridate")
library(ggplot2)
#library(dplyr)

# knitr: Suppress code/messages/warnings 
#        Set default plot options and center them
knitr::opts_chunk$set(fig.width=9,fig.height=5,fig.path='Figs/',
                      fig.align='center',tidy=TRUE,
                      echo=FALSE,warning=FALSE,message=FALSE)

```

```{r echo=FALSE, loadData}
# Load the Data
loanData <- read.csv('prosperLoanData.csv')
```

Prosper Marketplace, Inc. is a San Francisco, California-based company in the peer-to-peer lending industry. Prosper Funding LLC, one of its subsidiaries, operates Prosper.com, a website where individuals can either invest in personal loans or request to borrow money. This data set contains 113,937 loans with 81 variables on each loan, including loan amount, borrower rate (or interest rate), current loan status, borrower income, borrower employment status, borrower credit history, and the latest payment information.

# Univariate Plots Section

### Data Structure
```{r echo=FALSE, structure}
str(loanData)

```

### Summary of the dataset
```{r echo=FALSE, summary}
summary(loanData)

```

### 1. Loan Status
Let’s take a look at what are the different loan status and which status are most borrower at?
```{r echo=FALSE, loanStatus}
table(loanData$LoanStatus)

```

**There are 12 different loan status:**

1. Cancelled
2. Chargedoff
3. Completed
4. Current
5. Defaulted
6. FinalPaymentInProgress
7. Past Due (>120 days)
8. Past Due (1-15 days)
9. Past Due (16-30 days)
10. Past Due (31-60 days)
11. Past Due (61-90 days)
12. Past Due (91-120 days)

* There are 6 different kinds of `PastDue` classified by different number of due days. Since I'm only concern about whether a loan is under high risk or not, I have combined all loans under 121 days of `PastDue` into one variable: `PastDue`.
```{r echo=FALSE, combinedPastDue}
#create list that I want to combine into PastDue variable
PastDue <- c("Past Due (1-15 days)",
             "Past Due (16-30 days)",
             "Past Due (31-60 days)",
             "Past Due (61-90 days)",
             "Past Due (91-120 days)")
```

* Based on the Prosper website, a borrower loan is charged-off when it reaches 121 days past due unless the borrower is otherwise notified by Prosper. There are 16 obseravtion with `Past Due(> 120 days)` so I will combined them into `Chargedoff` variable.
```{r echo=FALSE, combinedChargedoff}
#create list that I want to combine into Chargedoff variable
Chargedoff <- c("Past Due (>120 days)")
```

* There are 205 observations under `FinalPaymentInProgress` and since it is almost equivalent to `Completed`, I will combined it with `Completed`. 
```{r echo=FALSE, combinedCompleted}
#create list that I want to combine into Completed variable
Completed <- c("FinalPaymentInProgress")
```

```{r echo=FALSE, updateChanges}
#change LoanStatus format to string
loanData$LoanStatus <- as.character(loanData$LoanStatus)
# start change each variable
loanData$LoanStatus[loanData$LoanStatus %in% Chargedoff] <- "Chargedoff"
loanData$LoanStatus[loanData$LoanStatus %in% PastDue] <- "PastDue"
loanData$LoanStatus[loanData$LoanStatus %in% Completed] <- "Completed"
loanData$LoanStatus <- factor(loanData$LoanStatus, 
                          levels = c("Completed",  "Current", 
                                     "PastDue", "Chargedoff", "Defaulted", "Cancelled"), 
                          ordered = T)
# print out the change result
table(loanData$LoanStatus)
```

The bar chart of the loan status:

```{r echo=FALSE, LoanStatusPlot, dpi=36, out.width="700px", out.height="700px"}
# plot Loan Status bar chart
ggplot(aes(x= LoanStatus, y = ..count../sum(..count..)), data = loanData) + 
  geom_bar() +
  xlab('Loan Status') +
  ylab('Proportion of Borrowers (%)') +  
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +
      ggtitle("Distribution of Loan Status")
```

**Observations:**

* 34% of the loans are completed.
* The proportion of current loan is about 50%, which means that we are not sure if half of the loan data are defaulted or completed.
* 15% of the loans are chargedoff and defaulted which means there is risk that the investor may loss money.


### 2. Occupation

What are the borrower's occupation?

```{r echo=FALSE, occupation}
table(loanData$Occupation)
```

The bar chart of the borrower's occupation:
```{r echo=FALSE, ocupation}
# plot Employment Status bar chart
ggplot(aes( x = Occupation ), data = subset(loanData, Occupation != '')) +
  geom_bar(fill = '#599ad3' ) +
  xlab('Occupation') +
  ylab('No. of Borrowers') +  
    theme(axis.text.x = element_text(angle = 90, hjust=0.95,vjust=0.2, size =10)) +
        ggtitle("Distribution of Loans by occupation of the borrower")
```

```{r echo=FALSE, occupationPercentage}
# get borrower occupation percentage
OccupationPercentage <- table(loanData$Occupation)
round(prop.table(addmargins(OccupationPercentage)),digits=2)
```

**Observations:**

* There are a total of 68 different occupations.
* Most of the borrower's occupation is `Others`(13%) and `Professional` (6%). Seems like most borrower did not list down their occupation.


### 3. Employment Status

What are the different employment status?

```{r echo=FALSE, employmentStatus}
table(loanData$EmploymentStatus)
```

**There are 9 different employment status:**

1. Empty
2. Employed
3. Full-time
4. Not available
5. Not employed
6. Other
7. Part-time
8. Retired
9. Self-employed

* There are 2,255 empty status and 5,347 `Not available` status. I will hide them for the plot since they are not required.
* `Other` and `Not available` seems similar so that will be combined as `Other`.
```{r echo=FALSE, employmentStatusClean}
# combine both not available with other
Other <- c("Not available", 
           "Other")

loanData$EmploymentStatus[loanData$EmploymentStatus %in% Other] <- "Other"

# remove empty and Not available employment status
cleanEmploymentStatus <- loanData[!(is.na(loanData$EmploymentStatus) | 
                                      loanData$EmploymentStatus == ""), ]
```

The bar chart of the employment status:
```{r echo=FALSE, employmentStatusPlot_Cleaned}
# plot Employment Status bar chart
ggplot(data=cleanEmploymentStatus, aes(x= EmploymentStatus), y = ..count../sum(..count..)) + 
  geom_bar() +
  xlab('Employment Status') +
  ylab('Proportion of Borrowers (%)') +  
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +
      ggtitle("Distribution of Borrower's Employment Status")
```

**Observations:**

* Quite a number of employment status are left empty (2,255).
* Most of the borrowers are with stable income. 63% of the borrowers are employed and 25% of the borrowers are employed full-time.
* 8.2% of the borrowers are with Other employment status.

### 4. Is Borrower Homeowner

Are the borrower mostly homeowner or not?

```{r echo=FALSE, isBorrowerHomeowner}
table(loanData$IsBorrowerHomeowner)
```

The bar chart of whether borrow is homeowener:
```{r echo=FALSE, IsHomeownerPlot, dpi=36, out.width="700px", out.height="700px"}
# plot Loan Status bar chart
ggplot(aes(x= IsBorrowerHomeowner, y = ..count../sum(..count..)), data = loanData) + 
  geom_bar() +
  xlab('Is Borrower Homeowner') +
  ylab('Proportion of Borrowers (%)') +  
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +
      ggtitle("Propotion of homeowner")
```

**Observations:**

* There is slightly more borrower who are homeowner (50.5%) than those who are not a homeowner (49.6%).


### 5. Income Range

What are the income range of the borrowers?

```{r echo=FALSE, incomeRange}
table(loanData$IncomeRange)

```

* `Not displayed` values will be hidden since it is not useful for our analysis.
```{r echo=FALSE, rearrangeOrder}
# remove empty and Not available employment status
cleanIncomeRange <- loanData[!(is.na(loanData$IncomeRange) | 
                                      loanData$IncomeRange == "" | 
                                 loanData$IncomeRange == "Not displayed"), ]
```

The bar chart on Income Range:
```{r echo=FALSE, IncomeRangePlot, dpi=36, out.width="1000px", out.height="700px"}
# plot Income Range bar chart
ggplot(aes(x= IncomeRange, y = ..count../sum(..count..)), data = cleanIncomeRange) + 
  geom_bar() +
  xlab('Income Range ($)') +
  ylab('Proportion of Borrowers (%)') +  
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +
      ggtitle("Distribution of Borrower's Income Range")
```

**Observations:**

* Seems like the number of borrowers who are `Not employed`(835) under `Income Range` is different from the number of borrowers who are `Not employed` under `Employment Status`(806).


### 6. Borrower State

Where are most borrower from?

```{r echo=FALSE, borrowerState}
table(loanData$BorrowerState)

```

```{r echo=FALSE, borrowerStatePlot}
# plot borrower state bar chart
ggplot(aes( x = BorrowerState ), data = subset(loanData, BorrowerState != '')) +
  geom_bar(fill = '#599ad3' ) +
  xlab('Borrower State') +
  ylab('No. of Borrowers') +  
  theme(axis.text.x = element_text(angle = 90, hjust = .5, vjust = 0.5)) +
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +

          ggtitle("Distribution of Loans by state of the borrower")
```

**Observations:**

* Most of the borrowers are from California (13.6%) followed by Texas (6.3%) and New York (6.21%).


### 7. Loan Term

Let’s take a look at the loan term to see how long borrowers usually opt for loan?

```{r echo=FALSE, loanTerm}
table(loanData$Term)

```


The bar chart on loan term:
```{r echo=FALSE, loanTermPlot, dpi=36, out.width="1000px", out.height="700px"}
# plot Income Range bar chart
ggplot(aes(x= (Term/12), y = ..count../sum(..count..)), data = loanData) + 
  geom_bar() +
  xlab('Loan Term (Year)') +
  ylab('Proportion of Borrowers (%)') +  
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +
        ggtitle("Distribution of Loan Term")
```

**Observations:**

* There are 3 types of loan term - 12 months, 36 months and 60 months.
* 77% of the borrowers opt for 3 years loan term.
* Only 1% of the borrowers opt for 1 year loan term. Seems like most borrower are not interested in short loan term.


### 8. Prosper Rating (Alpha)

Let’s take a look at the prosper rating (Alpha)

```{r echo=FALSE, prosperRating}
table(loanData$ProsperRating..Alpha.)

```

```{r echo=FALSE, ProsperRating}
# plot Loan Status bar chart
loanData$ProsperRating <- factor(loanData$ProsperRating..Alpha, 
                          levels = c('AA', 'A', 'B', 'C', 'D', 'E', 'HR', 'NA'), 
                          ordered = T)

ggplot(aes(x= ProsperRating, y = ..count../sum(..count..)), data = loanData) + 
  geom_bar() +
  xlab('Prosper Rating') +
  ylab('Proportion of Borrowers (%)') +  
  geom_text(stat='count', 
            aes(label= paste(round(..count../sum(..count..)*100, 2),"%"))
            , vjust=-0.5) +
      ggtitle("Distribution of Prosper Rating")
```

**Observations:**

* There is huge percentage of prosper rating as NA (25.53%). 
* The highest prosper rating is C with 16.1% followed by B(13.68%) and A(12.77%).

### 9. Listing Creation Date

Which month or year had the most loan?

```{r echo=FALSE, loanMonth}
loanData$ListingCreationMonth <- as.factor(lubridate::month(loanData$ListingCreationDate))
loanData$ListingCreationYear <- as.factor(lubridate::year(loanData$ListingCreationDate))

# create Loan listing plot by Month
ggplot(aes(x = ListingCreationMonth), data = loanData) +
  geom_bar(fill = '#599ad3', color = 'black') + 
  xlab('Loan Listing Creation (Month)') +
  ylab('No. of Borrowers') +  
    scale_x_discrete(breaks = seq(1,12,1)) + 
      geom_text( stat = 'count', aes(label = ..count..), size = 3, vjust = -1) +
        ggtitle("Loan Creation by Month")
```

**Observations:**

* March (8,032) & April (7,661) have relatively lesser loan.
* January (11,214) had the highest number of loan.

```{r echo=FALSE, loanYearMonth}
# create Loan listing plot by Year & Month
ggplot(aes(x = ListingCreationMonth), data = loanData) +
  geom_bar(fill = '#599ad3', color = 'black') + 
  xlab('Loan Listing Creation (Year & Month)') +
  ylab('No. of Borrowers') +  
      facet_wrap(~ ListingCreationYear, ncol = 2 ) +
      scale_x_discrete(breaks = seq(1,12,1)) + 
        scale_y_sqrt() +
          ggtitle("Loan Creation by Year & Month")
```

**Observations:**

* Increase in the loans as year pass by except in Oct-2018 to 2019.

```{r echo=FALSE, loanYear}
# create Loan listing plot by Year
ggplot(aes(x = ListingCreationYear), data = loanData) +
  geom_bar(fill = '#599ad3', color = 'black') + 
  xlab('Loan Listing Creation (Year)') +
  ylab('No. of Borrowers') +  
      geom_text( stat = 'count', aes(label = ..count..), size = 3, vjust = -1) +
          ggtitle("Loan Creation by Year")
```

**Observations:**

* Sudden decrease of number of loan in 2009 (2,206).
* After a Google search was done, the decrease of number of loan in 2009 was due to the Prosper Quiet Period from October 15, 2008 until July 13, 2009. They closed down the borrower side of the business as well as the investor side. No new loans were issued during the Prosper quiet period. (Reference: https://www.lendacademy.com/a-look-back-at-the-lending-club-and-prosper-quiet-periods/)


# Univariate Analysis

### What is the structure of your dataset?

Our dataset consists of 81 variables, with 11,3937 observations.

### What is/are the main feature(s) of interest in your dataset?

Borrower: `Prosper Score` and `Proser Rating` seems to be the main indicators of a quality of borrowers.

Lender: `Lender Yield` seems to the most important factor for investor.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Properties about borrower such as `Employment Status`, `Income Range`, `Loan Term`.

### Did you create any new variables from existing variables in the dataset?

Yes I created some new variables and factorized a few others:

* `LoanStatus`: clean up the 12 different loan status into 6.
* `EmploymentStatus`: clean up the 9 different employment status into 9.
* Splited `ListingCreationDate` into Year and month for analysis

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

There are a number of empty fields and NA fields which create spikes in the plots:

* `EmploymentStatus` contains 2,255 empty blanks and 5,347 NA.
* `IncomeRange` contains a number of `Not displayed` values.
* 26% of `ProsperRating` is NA.

Had remove the blank and NA which are not required information for the plots creation.

# Bivariate Plots Section

### 1. Lender Yield vs Prosper Rating

Is there a relation between lender yield and prosper rating?

```{r echo=FALSE, ProsperRatingvsLenderYield}
loanData$ProsperRating..Alpha. <- factor(loanData$ProsperRating..Alpha., levels = c("AA", "A", "B", "C", "D", "E", "HR") )
# create boxplot - Lender Yield and Prosper Rating
ggplot(aes( x = ProsperRating..Alpha., y = LenderYield ), data = subset(loanData, ProsperRating..Alpha. != "")) +
  geom_boxplot() + 
  xlab('Prosper Rating') +
  ylab('Lender Yield (bps)') +  
      ggtitle("Lender Yield vs Prosper Rating")
```

**Observation:**

* There's a clear relation between the two, the higher the lender yield, the lower is the Prosper Rating.

### 2. Borrower Rate vs Prosper Rating

Is there a relation between borrower rate and prosper rating?

```{r echo=FALSE, borrowerRatevsProsperRating}
# create boxplot - Borrower Rate and Prosper Rating
ggplot(aes(x = ProsperRating..Alpha., y = BorrowerRate), data = subset(loanData, ProsperRating..Alpha. != "")) +
  geom_boxplot() +
  xlab('Prosper Rating') +
  ylab('Borrower Rate (bps)') +  
      ggtitle("Borrower Rate vs Prosper Rating")
```

**Observation:**

* The borrower rate is quite similar to the Lender Yield. The relationship observed here is also similar to the lender yield.


### 3. Loan Orignal Amount vs Prosper Rating

Is there a relation between loan original amount and prosper rating?

```{r echo=FALSE, loanAmtvsProsperRating}
# create boxplot - Loan Amount and Prosper Rating
ggplot(aes( x = ProsperRating..Alpha., y = LoanOriginalAmount ), data = subset(loanData, ProsperRating..Alpha. != "")) +
  geom_boxplot()  + 
  xlab('Prosper Rating') +
  ylab('Loan Amount ($)') +  
    coord_cartesian(ylim = c(0,20000)) + 
      ggtitle("Loan Orignal Amount vs Prosper Rating")
```

**Observation:**

* The Median Loan Amount appears to decrease as the prosper rating gets poorer.

### 4. Borrower Monthly Income vs Prosper Rating

Is there a relation between monthly income and prosper rating?

```{r echo=FALSE, borrowerIncomevsProsperRating}
# create boxplot - borrower monthly income and prosper rating
ggplot(aes(x = ProsperRating..Alpha., y = StatedMonthlyIncome), data = subset(loanData, ProsperRating..Alpha. != "")) +
  geom_boxplot() +
  xlab('Prosper Rating') +
  ylab('Borrower Monthly Income ($)') +  
     coord_cartesian(ylim = c(0,10000)) +
    ggtitle("Monthly Income of Borrower vs Prosper Rating")
```

**Observation:**

* The Monthly Income of the borrower shows a slight decrease as the rating gets poorer.

### 5. Monthly Income of Borrower vs Original Amount of the Loan

Is there a relation between monthly income and loan amount?

```{r echo=FALSE, monthlyIncomevsOriginalAmt}
# create scatterplot - borrower monthly income and loan amount
ggplot(aes(y = StatedMonthlyIncome, x = LoanOriginalAmount), data = subset(loanData, !is.na(StatedMonthlyIncome))) +
  geom_point( alpha = 1/20, position = position_jitter(h = 0) ) + 
      coord_cartesian(ylim = c(0, 50000)) +
      xlab('Loan Amount ($)') +
      ylab('Monthly Income ($)') +  
            geom_smooth() +
            ggtitle("Monthly Income of Borrower vs Original Amount of the Loan")
```

**Observation:**

* As the Loan Amount increases, the median monthly income also increases. Another interesting trend is that the Loan amount values appear to be discrete as the Loan amounts are usually round figures rather than random values.

### 6. Borrower Rate vs Status of Loan

Is there a relation between borrower rate and loan status?

```{r echo=FALSE, borrowerRatevsStatusOfLoan}
# create boxplot - borrower rate and loan status
ggplot(aes( x = LoanStatus, y = BorrowerRate ), data = loanData) +
  geom_boxplot() + 
    theme(axis.text.x = element_text(angle = 90, hjust=0.95,vjust=0.2, size =10)) +
      xlab('Loan Status') +
      ylab('Borrower Rate (bps)') +  
        ggtitle("Borrower Rate vs Status of Loan")
```

**Observation:**

* The median of Borrower Rate for defaulted and chargedoff loans appear to be a bit higher than the others.

### 7. Occupation of Borrower vs Original Amount of Loan

Is there a relation between occupation and loan amount?

```{r echo=FALSE, occupationVsLoanAmt, fig.width = 12, fig.height = 8}
# create boxplot - occupation and loan amount
ggplot(aes( x = Occupation, y = LoanOriginalAmount ), data = loanData) +
  geom_boxplot() + 
      xlab('Occupation') +
      ylab('Loan Amount ($)') +  
      theme(axis.text.x = element_text(angle = 90, hjust=0.95,vjust=0.2, size =7)) +
      ggtitle("Occupation of Borrower vs Original Amount of Loan") +
  coord_flip(ylim = c(0, 20000))+
  theme(axis.text.y = element_text( size =9))
```

**Observation:**

* This shows the variations in the amount taken as loan by people from different occupation.
* Occupation such as Judge have quite a range difference in loan amount.

### 8. Occupation of Borrower vs his Monthly Income

Is there a relation between occupation and monthly income?

```{r echo=FALSE, OccupationVsMonthlyIncome, fig.width = 12, fig.height = 8}
# create boxplot - occupation and monthly income
ggplot(aes( x = Occupation, y = StatedMonthlyIncome ), data = loanData) +
  geom_boxplot() + 
      xlab('Occupation') +
      ylab('Monthly Income ($)') +  
      theme(axis.text.x = element_text(angle = 90, hjust=0.95,vjust=0.2, size =7)) + 
      coord_cartesian(ylim = c(0, 20000)) +
        ggtitle("Occupation of Borrower vs his Monthly Income")+
  coord_flip(ylim = c(0, 20000))+
  theme(axis.text.y = element_text( size =9))
```

**Observation:**

* The variation in monthly income for different occupations.
* Occupation such as Doctor have quite a high range in Income.


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

There is clear relation between `Prosper Rating`, `Lender Yield`, `Borrower Rate` and `Loan Amount`.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

`Borrower Rate` and `Lender Yield` have a linear relationship.

### What was the strongest relationship you found?

`Borrower rate` and `Lender Yield`.

# Multivariate Plots Section

### 1. Lender Yield vs Loan Amount vs Prosper Rating
```{r echo=FALSE, lenderYieldvsLoanAmtvsProsperRating}
# create scatterplot - lender yield, loan amoount and prosper rating
ggplot(aes(x = LenderYield, y = LoanOriginalAmount, color = ProsperRating..Alpha.),
       data = subset(loanData, !is.na(ProsperRating..Alpha.) & !is.na(LoanOriginalAmount) )) +
  geom_point(alpha = 1/2, position = position_jitter(h=0) ) +
      xlab('Lender Yield (bps)') +
      ylab('Loan Amount ($)') +  
      coord_cartesian(ylim = c(0, 20000)) + 
      ggtitle("Lender Yield vs Loan Amount vs Prosper Rating") +
  scale_color_brewer(palette = "RdYlBu", name="Prosper\nRating",
                     direction=-1)+
  theme(panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = "darkgrey"))  
  guides(color=guide_legend(override.aes = list(alpha = 1)))
```

**Observation:**

* As the Lender Yield increases, the Prosper Rating gets poorer.


### 2. Monthly Income vs Loan Amount vs Prosper Rating
```{r echo=FALSE, monthlyIncomeVsLoanAmountVsProsperRating}
# create scatterplot - monthly income, loan amount and prosper
ggplot(aes(x = StatedMonthlyIncome, y = LoanOriginalAmount, color = ProsperRating..Alpha.),
       data = subset(loanData, !is.na(ProsperRating..Alpha.) )) +
  geom_point(alpha = 1/2, position = position_jitter(h=0)) +
      xlab('Monthly Income ($)') +
      ylab('Loan Amount ($)') +  
      coord_cartesian(xlim = c(0, 20000)) +
      ggtitle("Monthly Income vs Loan Amount vs Prosper Rating")+
  scale_color_brewer(palette = "RdYlBu", name="Prosper\nRating",
                     direction=-1)+
  theme(panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = "darkgrey"))  
  guides(color=guide_legend(override.aes = list(alpha = 1)))
```

**Observation:**

* As the Loan Amount increases, the Prosper Rating gets better.

### 3. Lender Yield vs Monthly Income vs Prosper Rating
```{r echo=FALSE, LenderYieldVsMonthlyIncomeVsProsperRating}
# create scatterplot - lender yield, monthly and prosper rating
ggplot(aes(x = LenderYield, y = StatedMonthlyIncome, color = ProsperRating..Alpha.),
       data = subset(loanData, ProsperRating..Alpha. != "" )) +
      xlab('Lender Yield (bps)') +
      ylab('Monthly Income ($)') +  
    geom_point(alpha = 1/2, position = position_jitter(h=0)) +
    coord_cartesian(ylim = c(0, 20000)) +    
      ggtitle("Lender Yield vs Monthly Income vs Prosper Rating")+
  scale_color_brewer(palette = "RdYlBu", name="Prosper\nRating",
                     direction=-1)+
  theme(panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = "darkgrey"))  
  guides(color=guide_legend(override.aes = list(alpha = 1)))
```

**Observation:**

* Borrower's monthly income does not seems to have any relation with the Prosper Rating.

### 4. Lender Yield vs Borrower Rate vs Prosper Rating
```{r echo=FALSE, LenderYieldVsBorrowerRateVsProsperRating}
# create scatterplot - lener yield, borrower rate and prosper rating
ggplot(aes(x = LenderYield, y = BorrowerRate, color = ProsperRating..Alpha.),
       data = subset(loanData, ProsperRating..Alpha. != "" )) +
      xlab('Lender Yield (bps)') +
      ylab('Borrower Rate (bps') +  
    geom_point(alpha = 1/2, position = position_jitter(h=0)) +
    ggtitle("Lender Yield vs Borrower Rate vs Prosper Rating") +
  scale_color_brewer(palette = "RdYlBu", name="Prosper\nRating",
                     direction=-1)
```
**Observation:**

* Borrower Rate and Lender Yield variables have a strongly positive linear relationship between them. Both rise above their means or fall below their means at the same time.
* As Borrower Rate and Lender Yield increases, the prosper rating gets poorer.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Based on the above analysis, I can see that there are relation between Lender yield, Borrower rate and Loan amount in predicting the Prosper Rating.

### Were there any interesting or surprising interactions between features?

`Borrower Rte` and `Lender Yield` have a strongly positive linear relationship.


------

# Final Plots and Summary

### Plot One: Sudden drop of loan in 2009
```{r echo=FALSE, Plot_One}
# create Loan listing plot by Year & Month
ggplot(aes(x = ListingCreationMonth), data = loanData) +
  geom_bar(fill = '#599ad3', color = 'black') + 
  xlab('Loan Listing Creation (Year & Month)') +
  ylab('No. of Borrowers') +  
      facet_wrap(~ ListingCreationYear, ncol = 2 ) +
      scale_x_discrete(breaks = seq(1,12,1)) + 
        scale_y_sqrt() +
          ggtitle("Loan Creation by Year & Month")
```

This plot is useful as it shows there was a sudden drop in loan during Oct-2008 to Jun-2009. It was after a Google search and I realized the decrease of number of loan in 2009 was due to the Prosper Quiet Period from October 15, 2008 until July 13, 2009. They closed down the borrower side of the business as well as the investor side. No new loans were issued during the Prosper quiet period. (Reference: https://www.lendacademy.com/a-look-back-at-the-lending-club-and-prosper-quiet-periods/)

### Plot Two: Relationship between Borrower Occupation and Monthly Income
```{r echo=FALSE, Plot_Three, fig.width = 12, fig.height = 8}
# create boxplot - occupation and monthly income
ggplot(aes( x = Occupation, y = StatedMonthlyIncome ), data = loanData) +
  geom_boxplot() + 
      xlab('Occupation') +
      ylab('Monthly Income ($)') +  
      theme(axis.text.x = element_text(angle = 90, hjust=0.95,vjust=0.2, size =7)) + 
      coord_cartesian(ylim = c(0, 20000)) +
        ggtitle("Occupation of Borrower vs his Monthly Income")+
  coord_flip(ylim = c(0, 20000))+
  theme(axis.text.y = element_text( size =9))
```

This plot is interesting as the above diagram tells alot of information. It shows the variation in borrower's monthly income for the different occupation. Occupation such as Doctor have quite a high range in monthly income. Most people had selected others and professionals as their occupation. Obviously the students have the lower range in monthly income.


### Plot Three: Relationship between Lender Yield, Loan Amount and Prosper Rating
```{r echo=FALSE, Plot_Two}
# create scatterplot - lender yield, loan amount and prosper
ggplot(aes(x = LenderYield, y = LoanOriginalAmount, color = ProsperRating..Alpha.),
       data = subset(loanData, !is.na(ProsperRating..Alpha.) & !is.na(LoanOriginalAmount) )) +
  geom_point(alpha = 1/2, position = position_jitter(h=0) ) +
      xlab('Lender Yield (bps)') +
      ylab('Loan Amount ($)') +  
      coord_cartesian(ylim = c(0, 20000)) + 
      ggtitle("Lender Yield vs Loan Amount vs Prosper Rating") +
  scale_color_brewer(palette = "RdYlBu", name="Prosper\nRating",
                     direction=-1)+
  theme(panel.grid.minor = element_blank(), 
        panel.background = element_rect(fill = "darkgrey"))  
  guides(color=guide_legend(override.aes = list(alpha = 1)))
```

The above plot is useful by showing the relationship between Lender Yield, Loan Amount and Prosper Rating. As the Lender Yield gets higher.


------

# Reflection

The Loan Data from Prosper was a challenging data set but a good and useful way for me to learn and experience doing exploratory data analysis with R. It was diffult due to large number of data and variables available. It took me some time to figure out which are the the useful variables and which are the ones that I want to focus on. 

This project made me to appreciate all the wonderful visualizations and ways to reveal underlining information hidden in a dataset. Th experience gained with this dataset definitely improved my EDA skills. I tried to keep all my visualization as simple as possible to make them easy to read.

The analysis could be enriched in future work with more analysis done on the other variables or look at the variables through the investors side or borrowers side. I also want to learn and explore mode different useful plots and graphs.